Draft Letter

### 2024-03-14 rawfiles/Draft_letter.qmd
replaced figure 3 with individual ages rather than 5 year steps

While the existence of centuries of mortality measurement (not mentioned in Gibson's article) does not directly address if New Zealand medical professionals are aware that best practice mortality calculations recognise that people of different ages die at different rates when using the publicly available Our World in Data graphs in media pieces, it does establish that there is an established methodology, not mentioned in Gibson's article, underpinning claims of New Zealand's negative excess, and that method has been used for centuries, world-wide, across multiple disciplines. We can check if New Zealand medical professionals are aware of this by asking them: *"the key point being any comparisons both within country (over time) or between countries should be age standardised to provide the most accurate picture possible (and of course be based on high quality and complete underlying deaths data which we can say NZ’s are!)"*(Ashley Bloomfield) [^6].

@@

![Comparison of ages of migration and death](md_figures/immigration.png){fig-alt="Young people migrate, old people die" fig-align="center"}

@@

![Even if all years are before COVID-19, the post-baseline years are above trend](md_figures/figure5.png){fig-alt="Aging increases deaths and the crude death rate" fig-align="center"} Of more surprise is that Gibson's results for 2020-22 less well approximate the crude mortality rate than Karlinsky & Kobak, since the crude mortality rate is the accepted method of comparing death rates while ignoring the effects of age. While there is not enough information in the paper to exactly replicate Gibson's method and achieve the same answers, part of the reason for the difference seems likely to be the unusual step of treating population as an additive contributor to deaths in his modification of the regression. But, compared to the differences with methods acknowledging age, such problems are minor.

>>>>

While the existence of centuries of mortality measurement (not mentioned in Gibson's article) does not directly address if New Zealand medical professionals are aware that best practice mortality calculations recognise that people of different ages die at different rates when using the publicly available Our World in Data graphs in media pieces, it does establish that there is an established methodology, not mentioned in Gibson's article, underpinning claims of New Zealand's negative excess, and that method has been used for centuries, world-wide, across multiple disciplines. We can check if New Zealand medical professionals are aware of this by asking them: *"the key point being any comparisons both within country (over time) or between countries should be age standardised to provide the most accurate picture possible (and of course be based on high quality and complete underlying deaths data which we can say NZ's are!)"*(Ashley Bloomfield) [^6].

@@

![Figure 3: Ages of migration and ages of death have little overlap](md_figures/figure3.png){fig-alt="Young people migrate, old people die" fig-align="center"}

@@

![Figure 4: Applying the same test to pre-covid years shows the same aging population effects](md_figures/figure5.png){fig-alt="Aging increases deaths and the crude death rate" fig-align="center"}

Of more surprise is that Gibson's results for 2020-22 less well approximate the crude mortality rate than Karlinsky & Kobak, since the crude mortality rate is the accepted method of comparing death rates while ignoring the effects of age. While there is not enough information in the paper to exactly replicate Gibson's method and achieve the same answers, part of the reason for the difference seems likely to be the unusual step of treating population as an additive contributor to deaths in his modification of the regression. But, compared to the differences with methods acknowledging age, such problems are minor.
